### 2015-12-10 paper/4-performance.tex
Introduction to memory usage section

\subsection{Memory Usage}
TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. This is one reason why memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.
The platform used for memory profiling has an i5-3317U Intel CPU and 8GiB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).
\subsubsection{Measure}
Adevs passes messages by value, dxex passes a pointer. The runtime effects of this choice have already been demonstrated in the Interconnect benchmark, so, in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user, this is the actual memory footprint of the application. It is important to note that, especially in the optimistic case, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[htb]
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con&dxex opt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
\end{table}
Since, in the conservative case, messages are passed by pa ointer, a GVT/LBTS implementation is required to organise the garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in the conservative case. 
Moreover, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to an increased memory pressure. Unlike conservative and sequential execution, memory usage in the optimistic case varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{PHold}
\begin{table}[htb]
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con &dxex opt\\ \hline
		40 & x & 37 & 61 & 682  \\ \hline
	\end{tabular}
	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
\end{table}
With only 10\% of all messages being inter-kernel, we expect conservative to have memory consumption near that of the single threaded implementation, since intra-kernel messages are reclaimable after each round. The counterintuitive high memory usage can be explained by conservative's stalled round behaviour which occurs whenever a kernel cannot advance (eit $==$ time). In such a round messages are sent out but the kernel does not execute any transitions until it has received all input from all influencing kernels. \\
With lookahead $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation. The resulting longer time intervals within which no new lbts is found delay memory deallocation.
Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.
Optimistic exhibits the expected high memory usage.
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead. 
\begin{table}
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con & dxex opt\\ \hline
		39 & 39 & 35 & 43 & 259 \\ \hline
	\end{tabular}
	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
\end{table}

>>>>

\subsection{Memory Usage}
Apart from simulation execution time, memory usage during simulation is also of great importance.
While execution time only becomes a problem if it takes way too long, coming short only a bit of memory can make simulation infeasible.
We therefore also investigate memory usage of different synchronization protocols.

We do not tackle the problem of states that become too large for a single machine to hold.
This problem can be mitigated by distribution over multiple machines, which neither dxex or adevs support.
Comparison of memory usage should thus only be seen in the context of which models are feasible to be simulated by the tool.

\subsubsection{Remarks}
Both dxex and adevs use tcmalloc as memory allocator.
Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (\textit{e.g.}, malloc and free).
Tcmalloc only gradually releases memory back to the OS, whereas our pools will not do so at all.
If memory has been allocated once, it is, at least from a performance point of view, better to keep that memory in the pool indefinitely.
Due to our motivation for memory usage analysis, we will only measure peak allocation.
Profiling is done using Valgrind's massif tool~\cite{Nethercote:2007:VFH:1273442.1250746}.

\subsubsection{Results}
\begin{figure}
    \includegraphics[width=\columnwidth]{fig/memory_voorlopig.eps}
    \caption{Memory usage results.}
    \label{fig:memory}
\end{figure}

Adevs passes messages by value, dxex passes a pointer. The runtime effects of this choice have already been demonstrated in the Interconnect benchmark, so, in this section, we measure memory usage in number of allocated pages combining text, stack and heap memory for the program profiled. For the OS and/or user, this is the actual memory footprint of the application. It is important to note that, especially in the optimistic case, not all this memory is always in use by the kernels. During simulation, the pools will generally not return deallocated memory to the OS, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[htb]
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con&dxex opt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
\end{table}
Since, in the conservative case, messages are passed by pa ointer, a GVT/LBTS implementation is required to organise the garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic's TimeWarp requires state/event saving, and its GVT algorithm is more complex (with a resulting higher latency) than the LBTS calculation in the conservative case. 
Moreover, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. Leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to an increased memory pressure. Unlike conservative and sequential execution, memory usage in the optimistic case varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{PHold}
\begin{table}[htb]
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con &dxex opt\\ \hline
		40 & x & 37 & 61 & 682  \\ \hline
	\end{tabular}
	\caption{Phold n 4 s 16 t1e6 r 10, unit MiB, 4 kernels (if parallel)}
\end{table}
With only 10\% of all messages being inter-kernel, we expect conservative to have memory consumption near that of the single threaded implementation, since intra-kernel messages are reclaimable after each round. The counterintuitive high memory usage can be explained by conservative's stalled round behaviour which occurs whenever a kernel cannot advance (eit $==$ time). In such a round messages are sent out but the kernel does not execute any transitions until it has received all input from all influencing kernels. \\
With lookahead $\epsilon$ this then leads to a high frequency of polling on shared null-times, which are used to determine lbts and thus garbage collection.
The lbts calculation will not wait until a new value is found, since this can create unwanted contention with the simulation. The resulting longer time intervals within which no new lbts is found delay memory deallocation.
Adevs' conservative fails to complete the benchmark under valgrind even with a significantly reduced load.
Optimistic exhibits the expected high memory usage.
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. Interconnect highlights dxex's lower performance due to message allocation overhead. 
\begin{table}
	\centering
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		adevs & adevs con &dxex &dxex con & dxex opt\\ \hline
		39 & 39 & 35 & 43 & 259 \\ \hline
	\end{tabular}
	\caption{Interconnect w 20 t5e5, unit MiB, 2 kernels (if parallel)}
\end{table}
